Left-Linear Rewriting in Adhesive Categories

### 2024-07-10 concur/final.tex
pass finale su concur
minori

\emph{Synopsis.}
% The paper has the following structure. 
In \S\ref{sec:ade} we recall the basic 
definitions of DPO rewriting for $\mathcal{M}$-adhesive categories. 
In \S\ref{sec:equi} we present the key novelties of our proposal,
introducing the standard notion of independence and 
a novel axiomatic notion of switchability, showing that the
former fails to imply the latter for left-linear rewriting systems.
We then define well-switching rewriting systems,
a subclass of left-linear rewriting systems,
and we prove that the classical results for linear rewriting systems
are recovered for such a class.
Finally, in \S\ref{conclusions} we outline directions for future work.
%
%The appendices
% collect additional data,
Appendix~\ref{app:ade} recalls basic results on adhesive categories and
Appendix~\ref{app:fill}
presents \emph{strong enforcing} rewriting systems, a class of systems
where the local Church-Rosser Theorem holds.
Appendix C of the
extended version provides the proofs
of our results.

\section{$\mathcal{M}$-adhesive categories and rewriting systems}\label{sec:ade}
This  section recalls the basic theory of \emph{$\mathcal{M}$-adhesive categories} \cite{azzi2019essence,ehrig2012,ehrig2014adhesive,lack2005adhesive,heindel2009category}.
%
%Throughout the paper
Given a category $\X$ we will not distinguish
notationally between $\X$ and its class of objects, so
``$X\in \X$'' means that $X$ is an object of $\X$.
We let $\mor(\X)$, $\mon(\X)$ and $\reg(\X)$ denote the class of
all arrows, monos and regular monos of $\X$, respectively.  Given an
integer $n\in \mathbb{Z}$, $[0,n]$ denotes the set of natural numbers
less than or equal to $n$; in particular, $[0,n]=\emptyset$ if
$n<0$.

@@

\noindent
\parbox{8cm}{
  \begin{definition}[Abstraction equivalence]
    Let  $\X$ be an $\mathcal{M}$-adhesive category and 
    $(\X, \R)$ a left-linear rewriting system. 
    An \emph{abstraction equivalence} between derivations $\der{D}$
    and $\der{D'}$ with the same length and
    $r(\der{D})=r(\der{D}')$ is a family of isomorphisms
    $\{\phi_X\}_{X\in \{G_i,D_i\}}$ such that the diagram on the right commutes
    for $i\in [0, \min(0, \lgh(\der{D})-1 )]$. We say that $\der{D}$ are $\der{D}'$ are \emph{abstraction
      equivalent}, written $\der{D}\equiv_a \der{D}'$, if there exists an abstraction equivalence between
    them.
  \end{definition}
}
\parbox{4cm}{\vspace{-1.35em}$\xymatrix@C=40pt{G'_i&D'_i \ar[r]^{g'_i}
    \ar@{>->}[l]_{f'_i}&G'_{i+1}\\ L_i \ar[u]^{m'_i} \ar[d]_{m_i}& K_i
    \ar[u]^{k'_i} \ar[d]_{k_i} \ar[r]^{r_i} \ar@{>->}[l]_{l_i}
    &R_i \ar[u]^{h'_i} \ar[d]_{h_i}\\G_i
    \ar@/_.45cm/[uu]_(.35){\phi_{G_i}}|\hole&D_i\ar@{>->}[l]^{f_i}\ar@/_.45cm/[uu]_(.35){\phi_{D_i}}|\hole
    \ar[r]_{g_i}&G_{i+1}\ar@/_.45cm/[uu]_{\phi_{G_{i+1}}}}$}

@@

\noindent 
\parbox{10cm}{
\begin{proposition}\label{prop:stab}Let $\X$ be a category and $\mathcal{A}$ a class of arrows stable under pullbacks. Suppose that, in the diagram aside, the whole rectangle is an $\mathcal{A}$-stable pushout and the right square a pullback. If the arrow $k$ is in $\mathcal{A}$ and it is a monomorphism,  then both squares are pushouts.
\end{proposition}}
\parbox{4cm}{\vspace{-.3cm}
	$\xymatrix{X \ar[d]_{a} \ar[r]^{f}& \ar[r]^{g} Y \ar[d]^{b}& Z \ar[d]^{c}\\ A \ar[r]_{h}& B \ar[r]_{k}& C}$}

@@

\noindent
\parbox{8.5cm}{
	\begin{lemma}\label{lem:radj}
		Let $f\colon X \to Y$ be an arrow in an $\mathcal{M}$-adhesive category $\X$ and suppose that the left square aside is a pushout while the  right one is a pullback, with $ m\colon M \rightarrowtail X$ and $n\colon N \rightarrowtail Y$ in $\mathcal{M}$.
		Then $n\leq k$ if and only if $p_2\leq m$.
	\end{lemma}}
\parbox{3cm}{\vspace{-2ex}
	$\xymatrix{M \ar@{>->}[d]_{m}\ar[r]^{h}& Q \ar@{>->}[d]^{k} & P \ar[r]^{p_1} \ar@{>->}[d]_{p_2}& N \ar@{>->}[d]^{n}\\X\ar[r]_{f}&Y &X \ar[r]_f&Y}$}

@@

\noindent
\parbox{10.4cm}{
	\begin{definition}Let $\mathcal{M}$ be a class of monos in a category $\X$, closed under composition, containing all isomorphisms and stable under pullbacks and pushouts. Suppose that the  diagram aside is given and the whole rectangle is a pushout. We say that $\X$ satisfies
		\parbox{14cm}{\begin{itemize}
				\item the \emph{$\mathcal{M}$-mixed decomposition} property if, whenever $k$ belongs to $\mathcal{M}$ and the right half of the previous diagram is a pullback, then the left one is a pushout;
				\item the \emph{$\mathcal{M}$-pushout decomposition} property if whenever $a, b$ and $c$ belongs to $\mathcal{M}$ and the right half of the diagram above is a pushout, then its left half is a pushout too.
			\end{itemize}}
	\end{definition}} \parbox{3cm}{\vspace{-2cm}$\xymatrix{X \ar[d]_{a} \ar[r]^{f}& \ar[r]^{g} Y \ar[d]^{b}& Z \ar[d]^{c}\\ A \ar[r]_{h}& B \ar[r]_{k}& C}$}\\

@@

\begin{definition} 
	Let $\X$ be an  $\mathcal{M}$-adhesive category, the pair $(\X, \mathcal{M})$ is of \emph{type $\mathbb{B}$} if
	\begin{enumerate}
		\item every arrow in $\mathcal{M}$ is a coproduct coprojection;
		\item $\X$ has all pushouts;
		\item $\X$ has strict initial objects and, for every $X$, the unique arrow $?_X\colon 0\to X $ belongs to $\mathcal{M}$;
		\item all pushouts are $\mathcal{M}$-stable.
	\end{enumerate}
\end{definition}

>>>>

\emph{Synopsis.}
% The paper has the following structure. 
In \S\ref{sec:ade} we recall the basic 
definitions of DPO rewriting for $\mathcal{M}$-adhesive categories. 
In \S\ref{sec:equi} we present the key novelties of our proposal,
introducing the standard notion of independence and 
a novel axiomatic notion of switchability, showing that the
former fails to imply the latter for left-linear rewriting systems.
We then define well-switching rewriting systems,
a subclass of left-linear rewriting systems,
and we prove that the classical results for linear rewriting systems
are recovered for such a class.
Finally, in \S\ref{conclusions} we outline directions for future work.
%
%The appendices
% collect additional data,
Appendix~\ref{app:ade} recalls basic results on adhesive categories and
Appendix~\ref{app:fill}
presents \emph{strong enforcing} rewriting systems, a class of systems
where the local Church-Rosser Theorem holds. Full proofs of our results are available in the extended version of the paper \cite{baldan2024full}.
\section{$\mathcal{M}$-adhesive categories and rewriting systems}\label{sec:ade}
This  section recalls the basic theory of \emph{$\mathcal{M}$-adhesive categories} \cite{azzi2019essence,ehrig2012,ehrig2014adhesive,lack2005adhesive,heindel2009category}.
%
%Throughout the paper
Given a category $\X$ we will not distinguish
notationally between $\X$ and its class of objects, so
``$X\in \X$'' means that $X$ is an object of $\X$.
We let $\mor(\X)$, $\mon(\X)$ and $\reg(\X)$ denote the class of
all arrows, monos and regular monos of $\X$, respectively.  Given an
integer $n\in \mathbb{Z}$, $[0,n]$ denotes the set of natural numbers
less than or equal to $n$; in particular, $[0,n]=\emptyset$ if
$n<0$.

@@

\noindent
\parbox{7.8cm}{
  \begin{definition}[Abstraction equivalence]
    Let  $\X$ be an $\mathcal{M}$-adhesive category and 
    $(\X, \R)$ a left-linear rewriting system. 
    An \emph{abstraction equivalence} between derivations $\der{D}$
    and $\der{D'}$ with the same length and
    $r(\der{D})=r(\der{D}')$ is a family of isomorphisms
    $\{\phi_X\}_{X\in \{G_i,D_i\}}$ such that the diagram on the right commutes
    for $i\in [0, \min(0, \lgh(\der{D})-1 )]$. We say that $\der{D}$ are $\der{D}'$ are \emph{abstraction
      equivalent}, written $\der{D}\equiv_a \der{D}'$, if there exists an abstraction equivalence between
    them.
  \end{definition}
}
\parbox{4.2cm}{\vspace{-1.35em}$\xymatrix@C=40pt{G'_i&D'_i \ar[r]^{g'_i}
    \ar@{>->}[l]_{f'_i}&G'_{i+1}\\ L_i \ar[u]^{m'_i} \ar[d]_{m_i}& K_i
    \ar[u]^{k'_i} \ar[d]_{k_i} \ar[r]^{r_i} \ar@{>->}[l]_{l_i}
    &R_i \ar[u]^{h'_i} \ar[d]_{h_i}\\G_i
    \ar@/_.45cm/[uu]_(.35){\phi_{G_i}}|\hole&D_i\ar@{>->}[l]^{f_i}\ar@/_.45cm/[uu]_(.35){\phi_{D_i}}|\hole
    \ar[r]_{g_i}&G_{i+1}\ar@/_.45cm/[uu]_{\phi_{G_{i+1}}}}$}

@@

\noindent 
\parbox{10cm}{
\begin{proposition}\label{prop:stab}Let $\X$ be a category and $\mathcal{A}$ a class of arrows stable under pullbacks. Suppose that, in the diagram aside, the whole rectangle is an $\mathcal{A}$-stable pushout and the right square a pullback. If the arrow $k$ is in $\mathcal{A}$ and it is a monomorphism,  then both squares are pushouts.
\end{proposition}}
\parbox{4cm}{\vspace{-.15cm}
	$\xymatrix{X \ar[d]_{a} \ar[r]^{f}& \ar[r]^{g} Y \ar[d]^{b}& Z \ar[d]^{c}\\ A \ar[r]_{h}& B \ar[r]_{k}& C}$}

@@

\noindent
\parbox{8.5cm}{
	\begin{lemma}\label{lem:radj}
		Let $f\colon X \to Y$ be an arrow in an $\mathcal{M}$-adhesive category $\X$ and suppose that the left square aside is a pushout while the  right one is a pullback, with $ m\colon M \rightarrowtail X$ and $n\colon N \rightarrowtail Y$ in $\mathcal{M}$.
		Then $n\leq k$ if and only if $p_2\leq m$.
	\end{lemma}}
\parbox{3cm}{%\vspace{-2ex}
	$\xymatrix{M \ar@{>->}[d]_{m}\ar[r]^{h}& Q \ar@{>->}[d]^{k} & P \ar[r]^{p_1} \ar@{>->}[d]_{p_2}& N \ar@{>->}[d]^{n}\\X\ar[r]_{f}&Y &X \ar[r]_f&Y}$}

@@

\noindent
\parbox{10.4cm}{
	\begin{definition}Let $\mathcal{M}$ be a class of monos in a category $\X$, closed under composition, containing all isomorphisms and stable under pullbacks and pushouts. Suppose that the  diagram aside is given and the whole rectangle is a pushout. We say that $\X$ satisfies
		\parbox{14cm}{\begin{itemize}
				\item the \emph{$\mathcal{M}$-mixed decomposition} property if, whenever $k$ belongs to $\mathcal{M}$ and the right half of the previous diagram is a pullback, then the left one is a pushout;
				\item the \emph{$\mathcal{M}$-pushout decomposition} property if whenever $a, b$ and $c$ belongs to $\mathcal{M}$ and the right half of the diagram above is a pushout, then its left half is a pushout too.
			\end{itemize}}
	\end{definition}} \parbox{3cm}{\vspace{-1.8cm}
	$\xymatrix{X \ar[d]_{a} \ar[r]^{f}& \ar[r]^{g} Y \ar[d]^{b}& Z \ar[d]^{c}\\ A \ar[r]_{h}& B \ar[r]_{k}& C}$}\\

@@

\begin{definition} 
	Let $\X$ be an  $\mathcal{M}$-adhesive category, the pair $(\X, \mathcal{M})$ is of \emph{type $\mathbb{B}$} if
	\begin{enumerate}
		\item every arrow in $\mathcal{M}$ is a coproduct coprojection;
		\item $\X$ has all pushouts;
		\item $\X$ has strict initial objects and for every $X$ the unique arrow $?_X\colon 0\to X $ belongs to $\mathcal{M}$;
		\item all pushouts are $\mathcal{M}$-stable.
	\end{enumerate}
\end{definition}
